The impact of environmental shocks on children's schooling in Bolivia

### 2020-05-01 paper_draft.Rmd
paper coding draft separate from the data appendix

title: "Data Appendix to "The impact of weather events on schooling outcomes""
author: "Katya Garcia-Israel"
output:
  html_document:
    toc: yes
    df_print: paged
  pdf_document:
    toc: yes
    number_sections: yes
---

@@

```{r set dfSummary css style, echo = F, include = F}
st_css()
```
# Appendix description
This data appendix details the steps of collecting, processing, and merging the data needed to create the complete data set showing information about schooling and weather events. It concludes with a discussion of patterns found in the data.

The datasets used directly by the final analysis are saved in `processed-data/` at the end of this file.

# Raw data
This section documents the datasets used in this analysis.

## Household Survey data
**Citation:** .  
Instituto Nacional de Estadistica. (2005). Encuesta de Hogares (2005) [SAV file]. Retrieved from https://www.ine.gob.bo/index.php/banco/base-de-datos-sociales

**Date Downloaded:** Downloaded September 2019.
**Filename(s):** raw_data/filename.csv 
**Unit of observation:** Unit of observation at the individual level, but each individual is identified mainly by the household id. 
**Dates covered:** 2005

### To obtain a copy

This household survey can be found on the Bolivian National Statistical Institute website, under the menu for 'Banco de Datos' and 'Base de datos de encuestas sociales.' Then the household surveys may be found for the years 2005-2018, excepting 2010.

The original format of this data was from the sav file, but I imported it into Stata before importing the Stata format into R.

@@

### Data import code and summary

```{r import household data 2005}
data_2005 <- read_dta("Raw data/seminar_paper_data/2005_hhsurvey.dta") %>% 
  rename(hh_id = folio,
        assist1 = s3_08,  
        reason = s3_09, 
        members = nm_y,
        sex = s1_02,
        age = s1_03,
        relation = s1_05,
        literacy1 = s3_01,
        register1 = s3_04,
        school_level = s3_05a,
        grade = s3_05b,
        repeated = s3_07,
        income = y_perc1A,
        educ = aoesc,
        job = s4_16cod,
        tasks = s4_16b,
        pov_line = z,
        extr_pov = z_ext)

@@

```{r export summary table 2005}
export_summary_table(dfSummary(hh_2005))

@@

```{r import household data 2006}
data_2006 <- read_dta("Raw data/seminar_paper_data/2006_hhsurvey.dta")  %>% 
  rename(hh_id = Folio,
        assist1 = s4_08,  
        reason = S4_09, 
        members = mhogar,
        sex = s1_02,
        age = s1_03,
        relation = s1_05,
        literacy1 = s4_01,
        register1 = s4_04,
        school_level = s4_05a,
        grade = s4_05b,
        repeated = s4_07,
        income = yhogpcf,
        urb_rur = Urb_Rur,
        depto = ID01,
        educ = añoe,
        job = s5_17cod,
        tasks = s5_16b,
        pov_line = z,
        extr_pov = zext)

@@

```{r exporting data table}
export_summary_table(dfSummary(hh_2006))
```

```{r import household data 2007}
data_2007 <- read_dta("Raw data/seminar_paper_data/2007_hhsurvey.dta")  %>% 
  rename(hh_id = folio,
        assist1 = s4_10,  
        reason = s4_11, 
        members = mhogar,
        sex = s1_03,
        age = s1_04,
        relation = s1_06,
        literacy1 = s4_01,
        register1 = s4_06,
        school_level = s4_07a,
        grade = s4_07b,
        repeated = s4_09,
        income = yhogpcf,
        voucher = s4_05,
        educ = a_oe,
        job = s5_16co,
        tasks = s5_16b,
        pov_line = z,
        extr_pov = zext)

@@

```{r import household data 2011}
data_2011 <- read_dta("Raw data/seminar_paper_data/2011_hhsurvey.dta")  %>% 
  rename(hh_id = folio,
        assist1 = s4_12,  
        reason = s4_13,
        sex = s1_03,
        age = s1_04,
        relation = s1_08,
        literacy1 = s4_01,
        register1 = s4_05,
        grade = s4_06b,
        repeated = s4_11a,
        income = yhogpc,
        school_level = s4_06a,
        urb_rur = area,
        breakfast = s4_07,
        voucher = s4_08,
        voucher_spent = s4_09,
        educ = e,
        job = cod_16a,
        tasks = s5_16b,
        pov_line = z,
        extr_pov = zext)

@@

```{r household data 2012}
data_2012 <- read_dta("Raw data/seminar_paper_data/2012_hhsurvey.dta")  %>% 
  rename(hh_id = folio,
        assist1 = s4_12,  
        reason = s4_13,
        sex = s1_03,
        age = s1_04,
        relation = s1_08,
        literacy1 = s4_01,
        register1 = s4_04,
        grade = s4_05b,
        repeated = s4_11a,
        income = yhogpc,
        school_level = s4_05a,
        urb_rur = area,
        depto = departamento,
        breakfast = s4_07,
        voucher = s4_08,
        voucher_spent = s4_09,
        educ = e,
        job = cods5_16a,
        tasks = s5_16b,
        pov_line = z,
        extr_pov = zext)

@@

```{r combine household datasets}
all_years <- hh_2005 %>% 
  rbind(hh_2005, hh_2006, hh_2007, hh_2011, hh_2012)

@@

## Flood data
**Citation:** GeoSinager. (2016). Mapa Recurrencias de Inundaciones Gestiones 2002-2012. [CSV file]. Retrieved from http://geosinager.defensacivil.gob.bo/maps/226

**Date Downloaded:** October 2019
**Filename(s):** raw_data/filename.csv 
**Unit of observation:** Municipality  
**Dates covered:** 2002-2012

### To obtain a copy

@@

- **dept:** department name 
- **A_2005:** how many floods occurred in 2005 in that location 
- **A_2006:** how many floods occurred in 2006 in that location 
- **A_2007:** how many floods occurred in 2007 in that location 
- **A_2008:** how many floods occurred in 2008 in that location 
- **A_2009:** how many floods occurred in 2009 in that location 
- **A_2010:** how many floods occurred in 2010 in that location 
- **A_2011:** how many floods occurred in 2011 in that location 
- **A_2012:** how many floods occurred in 2012 in that location

### Data import code and summary
```{r import flood data}
flood <- read_csv(file.path("Raw data", "recurrencia_inundacion2002_2012bol.csv")) %>% 
  rename(dept = NOM_DEP)

@@

export_summary_table(dfSummary(flood_summary))
```
## Drought data
**Citation:** GeoSinager. (2016). Mapa Recurrencias de Sequias Gestiones 2002-2012. [CSV file]. Retrieved from http://geosinager.defensacivil.gob.bo/maps/225

**Date Downloaded:** October 2019.  
**Filename(s):** Raw data/droughts.csv 
**Unit of observation:** The unit of observation is the municipality 
**Dates covered:** 2002-2012

### To obtain a copy

@@

- **dept:** department name 
- **A_2005:** how many droughts occurred in 2005 in that location 
- **A_2006:** how many droughts occurred in 2006 in that location 
- **A_2007:** how many droughts occurred in 2007 in that location 
- **A_2008:** how many droughts occurred in 2008 in that location 
- **A_2009:** how many droughts occurred in 2009 in that location 
- **A_2010:** how many droughts occurred in 2010 in that location 
- **A_2011:** how many droughts occurred in 2011 in that location 
- **A_2012:** how many droughts occurred in 2012 in that location

@@

``` {r merging datasets}
total_data<- all_years %>% 
  left_join(flood_summary) %>%
  left_join(drought_summary) %>%
  mutate(child = case_when(relation == 3 ~ 1,
                           relation != 3 ~ 0),
         grandchild = case_when(relation == 8 ~ 1,
                                relation != 8 ~ 0),
         pov_diff = income - pov_line,
         pov = case_when(pov_diff > 0 ~ 0,
                         pov_diff < 0 ~ 1),
         extr_diff = income - extr_pov,
         extr = case_when (extr_diff > 0 ~ 0,
                           extr_diff < 0 ~ 1),
                  work_reason = case_when(reason == "work" ~ 1,
                                 reason != "work" ~ 0),
         money_reason = case_when(reason == "money" ~ 1, 
                                  reason != "money" ~ 0))

@@

# Analysis Variables

@@

```{r creating head of household subset}

@@

```{r creating children subset}

@@

```{r combining household head characteristics with children dataset}
hh_head_2 <- hh_head %>%
  select(female, age, literacy, educ, job, tasks, pov, extr, hh_id) %>%
  rename (female_hhhead = female,
          hhhead_age = age,
          hhhead_literacy = literacy,
          hhhead_educ = educ,
          hhhead_job = job,
          hhhead_tasks = tasks)

@@

{r creating summary statistics}
export_summary_table(dfSummary(total_subset))

install.packages("devtools")
devtools::install_github("ewenharrison/finalfit")
library(finalfit)
library(dplyr)

```{r regression variable specifications}

reg_data <- total_subset %>%
  mutate(year_factor = as.factor(year),
         income2 = case_when(income == 0 ~ 1), 
         lny = log(income2))

@@

```{r regressions total sample}

@@

#model_simple <- lm(assist ~ floods + droughts + year, data = reg_data)
#summary(model_simple)
model_assist <- lm(assist ~ floods + droughts + age + female + lny + child + grandchild + num_children + hhhead_educ + female_hhhead + rural + year_factor, data = reg_data)
summary(model_assist)
model_register <- lm(register ~ floods + droughts + age + female + lny + child + grandchild + num_children + hhhead_educ + female_hhhead + rural + year_factor, data = reg_data)
summary(model_register)
model_reason_t <- (lm(work_reason ~ floods + droughts + age + female + lny + child + grandchild + num_children + hhhead_educ + female_hhhead + year_factor, data = reg_data))
summary(model_reason_t)

@@

```{r generate regression table, echo = F, results = "asis"}
stargazer::stargazer(list(model_assist, model_register), 
               title = "Full sample regression results", 
               label = "fullResults",
               header = F,
               keep = c("floods", "droughts"))
```
```{r probit regression total sample}

probitMod_assist <- glm(assist ~ floods + droughts + age + female + lny + child + grandchild + num_children + hhhead_educ + female_hhhead + year_factor, data = reg_data, family=binomial(link="probit"))

probitMod_register <- glm(register ~ floods + droughts + age + female + lny + child + grandchild + num_children + hhhead_educ + female_hhhead + year_factor, data = reg_data, family=binomial(link="probit"))

probitMod_work <- glm(work_reason ~ floods + droughts + age + female + lny + child + grandchild + num_children + hhhead_educ + female_hhhead + year_factor, data = reg_data, family=binomial(link="probit"))

@@

```{r regression rural sample}

rural <- subset(reg_data, rural == 1, select = c("hh_id", "id", "age", "female", "literacy", "dept", "rural", "income", "assist", "register", "floods", "droughts", "year", "child", "grandchild", "num_children", "female_hhhead", "hhhead_literacy", "hhhead_educ", "year_factor", "work_reason"))

model_assist_r <- (lm(assist ~ floods + droughts + age + female + income + child + grandchild + num_children + hhhead_educ + female_hhhead + year_factor, data = rural))
summary(model_assist_r)
model_register_r <- lm(register ~ floods + droughts + age + female + income + child + grandchild + num_children + hhhead_educ + female_hhhead + year_factor, data = rural)
summary(model_register_r)
model_reason <- (lm(work_reason ~ floods + droughts + age + female + income + child + grandchild + num_children + hhhead_educ + female_hhhead + year_factor, data = rural))
summary(model_reason)

@@

```{r regression urban sample}

urban <- subset(reg_data, rural == 0, select = c("hh_id", "id", "age", "female", "literacy", "dept", "rural", "income", "assist", "register", "floods", "droughts", "year", "child", "grandchild", "num_children", "female_hhhead", "hhhead_literacy", "hhhead_educ", "year_factor"))

model_assist_u <- lm(assist ~ floods + droughts + age + female + income + child + grandchild + num_children + hhhead_educ + female_hhhead + year_factor, data = reg_data)
summary(model_assist_u)
model_register_u <- lm(register ~ floods + droughts + age + female + income + child + grandchild + num_children + hhhead_educ + female_hhhead + year_factor, data = reg_data)
summary(model_register_u)
```

>>>>

title: ""The impact of environmental shocks on children's schooling in Bolivia""
author: "Katya Garcia-Israel"
output:
  html_document:
    toc: yes
    df_print: paged
  pdf_document:
    toc: yes
    number_sections: yes
---

@@

```{r set dfSummary css style, echo = F, include = F}
st_css()
```
# Introduction

# Literature Review

# Data and Methods

## Data

The household data was taken from the social surveys database of the Bolivia National Statistical Institute, using household surveys from the years 2005-2007, 2011-2012.
Instituto Nacional de Estadistica. (2005). Encuesta de Hogares (2005) [SAV file]. Retrieved from https://www.ine.gob.bo/index.php/banco/base-de-datos-sociales

The unit of observation at the individual level, but each individual is identified mainly by the household id.

@@

```{r import household data 2005, echo = F, message = F}
data_2005 <- read_dta("Raw data/seminar_paper_data/2005_hhsurvey.dta") %>% 
  rename(hh_id = folio,
        assist1 = s3_08,  
        reason = s3_09, 
        members = nm_y,
        sex = s1_02,
        age = s1_03,
        relation = s1_05,
        literacy1 = s3_01,
        register1 = s3_04,
        school_level = s3_05a,
        grade = s3_05b,
        repeated = s3_07,
        income = y_perc1A,
        educ = aoesc,
        job = s4_16cod,
        tasks = s4_16b,
        pov_line = z,
        extr_pov = z_ext)

@@

```{r export summary table 2005, echo = F, message = F}
export_summary_table(dfSummary(hh_2005))

@@

```{r import household data 2006, echo = F, message = F}
data_2006 <- read_dta("Raw data/seminar_paper_data/2006_hhsurvey.dta")  %>% 
  rename(hh_id = Folio,
        assist1 = s4_08,  
        reason = S4_09, 
        members = mhogar,
        sex = s1_02,
        age = s1_03,
        relation = s1_05,
        literacy1 = s4_01,
        register1 = s4_04,
        school_level = s4_05a,
        grade = s4_05b,
        repeated = s4_07,
        income = yhogpcf,
        urb_rur = Urb_Rur,
        depto = ID01,
        educ = añoe,
        job = s5_17cod,
        tasks = s5_16b,
        pov_line = z,
        extr_pov = zext)

@@

```{r exporting data table, echo = F, message = F}
export_summary_table(dfSummary(hh_2006))
```

```{r import household data 2007, echo = F, message = F}
data_2007 <- read_dta("Raw data/seminar_paper_data/2007_hhsurvey.dta")  %>% 
  rename(hh_id = folio,
        assist1 = s4_10,  
        reason = s4_11, 
        members = mhogar,
        sex = s1_03,
        age = s1_04,
        relation = s1_06,
        literacy1 = s4_01,
        register1 = s4_06,
        school_level = s4_07a,
        grade = s4_07b,
        repeated = s4_09,
        income = yhogpcf,
        voucher = s4_05,
        educ = a_oe,
        job = s5_16co,
        tasks = s5_16b,
        pov_line = z,
        extr_pov = zext)

@@

```{r import household data 2011, echo = F, message = F}
data_2011 <- read_dta("Raw data/seminar_paper_data/2011_hhsurvey.dta")  %>% 
  rename(hh_id = folio,
        assist1 = s4_12,  
        reason = s4_13,
        sex = s1_03,
        age = s1_04,
        relation = s1_08,
        literacy1 = s4_01,
        register1 = s4_05,
        grade = s4_06b,
        repeated = s4_11a,
        income = yhogpc,
        school_level = s4_06a,
        urb_rur = area,
        breakfast = s4_07,
        voucher = s4_08,
        voucher_spent = s4_09,
        educ = e,
        job = cod_16a,
        tasks = s5_16b,
        pov_line = z,
        extr_pov = zext)

@@

```{r household data 2012, echo = F, message = F}
data_2012 <- read_dta("Raw data/seminar_paper_data/2012_hhsurvey.dta")  %>% 
  rename(hh_id = folio,
        assist1 = s4_12,  
        reason = s4_13,
        sex = s1_03,
        age = s1_04,
        relation = s1_08,
        literacy1 = s4_01,
        register1 = s4_04,
        grade = s4_05b,
        repeated = s4_11a,
        income = yhogpc,
        school_level = s4_05a,
        urb_rur = area,
        depto = departamento,
        breakfast = s4_07,
        voucher = s4_08,
        voucher_spent = s4_09,
        educ = e,
        job = cods5_16a,
        tasks = s5_16b,
        pov_line = z,
        extr_pov = zext)

@@

```{r combine household datasets, echo = F, message = F}
all_years <- hh_2005 %>% 
  rbind(hh_2005, hh_2006, hh_2007, hh_2011, hh_2012)

@@

The flood data is from the years 2005-2012. The unit of observation is municipality, but the data is later combined to make the unit of observation at the department level to make it comparable with the household survey data.

@@

```{r import flood data, echo = F, message = F}
flood <- read_csv(file.path("Raw data", "recurrencia_inundacion2002_2012bol.csv")) %>% 
  rename(dept = NOM_DEP)

@@

export_summary_table(dfSummary(flood_summary))
```
The drought data also spans the years 2005-2012 and shows the frequency of droughts in each municipality by year.

@@

@@

``` {r merging datasets, echo = F, message = F}
total_data<- all_years %>% 
  left_join(flood_summary) %>%
  left_join(drought_summary) %>%
  mutate(child = case_when(relation == 3 ~ 1,
                           relation != 3 ~ 0),
         grandchild = case_when(relation == 8 ~ 1,
                                relation != 8 ~ 0),
         pov_diff = income - pov_line,
         pov = case_when(pov_diff > 0 ~ 0,
                         pov_diff < 0 ~ 1),
         extr_diff = income - extr_pov,
         extr = case_when (extr_diff > 0 ~ 0,
                           extr_diff < 0 ~ 1),
                  work_reason = case_when(reason == "work" ~ 1,
                                 reason != "work" ~ 0),
         money_reason = case_when(reason == "money" ~ 1, 
                                  reason != "money" ~ 0))

@@

## Analysis Variables

@@

```{r creating head of household subset, echo = F, message = F}

@@

```{r creating children subset, echo = F, message = F}

@@

```{r combining household head characteristics with children dataset, echo = F, message = F}
hh_head_2 <- hh_head %>%
  select(female, age, literacy, educ, job, tasks, pov, extr, hh_id) %>%
  rename (female_hhhead = female,
          hhhead_age = age,
          hhhead_literacy = literacy,
          hhhead_educ = educ,
          hhhead_job = job,
          hhhead_tasks = tasks)

@@

```
```{r summary statistic tables with total_subset data, echo = F, message = F}
#creating a table with the main explanatory variables

#creating a table dividing the characteristics by urban/rural area
```
#Empirical Model

```{r regression variable specifications, echo = F, message = F}

reg_data <- total_subset %>%
  mutate(year_factor = as.factor(year),
         dept_factor = as.factor(dept),
         income2 = case_when(income == 0 ~ 1), 
         lny = log(income2))

@@

```{r linear regressions total sample, echo = F, message = F}

@@

#model_simple <- lm(assist ~ floods + droughts + year, data = reg_data)
#summary(model_simple)
model_assist <- lm(assist ~ floods + droughts + age + female + lny + child + grandchild + num_children + hhhead_educ + female_hhhead + rural + dept_factor + year_factor, data = reg_data)
summary(model_assist)
model_register <- lm(register ~ floods + droughts + age + female + lny + child + grandchild + num_children + hhhead_educ + female_hhhead + rural + dept_factor + year_factor, data = reg_data)
summary(model_register)
model_reason_t <- (lm(work_reason ~ floods + droughts + age + female + lny + child + grandchild + num_children + hhhead_educ + female_hhhead + dept_factor + year_factor, data = reg_data))
summary(model_reason_t)

@@

```{r generate regression table, echo = F, results = "asis"}
stargazer::stargazer(list(model_assist, model_register), 
               title = "Full sample regression results", 
               label = "fullResults",
               header = F,
               keep = c("floods", "droughts"))
```
```{r probit regression total sample, echo = F, message = F}

probitMod_assist <- glm(assist ~ floods + droughts + age + female + lny + child + grandchild + num_children + hhhead_educ + female_hhhead + dept_factor + year_factor, data = reg_data, family=binomial(link="probit"))

probitMod_register <- glm(register ~ floods + droughts + age + female + lny + child + grandchild + num_children + hhhead_educ + female_hhhead + dept_factor + year_factor, data = reg_data, family=binomial(link="probit"))

probitMod_work <- glm(work_reason ~ floods + droughts + age + female + lny + child + grandchild + num_children + hhhead_educ + female_hhhead + dept_factor + year_factor, data = reg_data, family=binomial(link="probit"))

@@

```{r regression rural sample, echo = F, message = F}

rural <- subset(reg_data, rural == 1, select = c("hh_id", "id", "age", "female", "literacy", "dept_factor", "rural", "income", "assist", "register", "floods", "droughts", "year", "child", "grandchild", "num_children", "female_hhhead", "hhhead_literacy", "hhhead_educ", "year_factor", "work_reason"))

model_assist_r <- (lm(assist ~ floods + droughts + age + female + income + child + grandchild + num_children + hhhead_educ + female_hhhead + dept_factor + year_factor, data = rural))
summary(model_assist_r)
model_register_r <- lm(register ~ floods + droughts + age + female + income + child + grandchild + num_children + hhhead_educ + female_hhhead + dept_factor + year_factor, data = rural)
summary(model_register_r)
model_reason <- (lm(work_reason ~ floods + droughts + age + female + income + child + grandchild + num_children + hhhead_educ + female_hhhead + dept_factor + year_factor, data = rural))
summary(model_reason)

@@

```{r linear regression urban sample, echo = F, message = F}

urban <- subset(reg_data, rural == 0, select = c("hh_id", "id", "age", "female", "literacy", "dept_factor", "rural", "income", "assist", "register", "floods", "droughts", "year", "child", "grandchild", "num_children", "female_hhhead", "hhhead_literacy", "hhhead_educ", "year_factor"))

model_assist_u <- lm(assist ~ floods + droughts + age + female + income + child + grandchild + num_children + hhhead_educ + female_hhhead + dept_factor + year_factor, data = reg_data)
summary(model_assist_u)
model_register_u <- lm(register ~ floods + droughts + age + female + income + child + grandchild + num_children + hhhead_educ + female_hhhead + dept_factor + year_factor, data = reg_data)
summary(model_register_u)
```
